---
title: "draftProject2"
author: "Shaya Engelman, Julia Ferris, Amanda Fox, and Jean Jimenez"
date: "2024-07-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

After completing the assignment, rework this as a Word Document to then submit.

## Data Prep

Import the data

```{r}
library(readr)
dataToModel <- read_csv("https://raw.githubusercontent.com/juliaDataScience-22/project2/main/studentDataToModel.csv")
View(dataToModel)

columnTypes <- read_csv("https://raw.githubusercontent.com/juliaDataScience-22/project2/main/dataColumnTypes.csv")
View(columnTypes)

dataToPredict <- read_csv("https://raw.githubusercontent.com/juliaDataScience-22/project2/main/studentEvaluationToPredict.csv")
View(dataToPredict)
```

Check for missing data

Remove duplicates?

Find errors, inconsistencies, outliers

Normalization and standardization if needed

Encode categorical variables

Merge datasets

Create new features from existing ones

Select relevant features

Split into test and train:



## Exploratory Data Analysis

## Model Selection / Evaluation

## Conclusions

## Sources